EP1162 Preliminary Analysis

### 2020-03-10 analysis/prelimAnalysis.Rmd
updated prelim analysis

## Prelim analysis
Below are some preliminary figures of the EP1162 data.  I tried different types of figures just to give us a few options to discuss when putting the final report together.  I was excited to start taking a look at the data!

@@

*Also - there may still be some adjustments made to two height measurements in the dataset.  It will be easy to re-run this analysis with an updated dataset if need  be!*

@@

dat %>% 
  left_join(plotDat,by="Plot") %>% 
  mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes

@@

*Could compare Q values between treatment units.*

### Trees per hectare by species in 2019
Let's look at species composition by treatment unit in 2019:

@@

dat %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data
   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
  filter(Year==2019) %>% # filter for 2019
  filter(Species%in%c("Sx","Bl")) %>% #filter for Sx and Bl
  group_by(BA.Target,Plot,Species,DBH.class) %>% 
  summarise(Spp.Plot=n()) %>% # summarise number of trees per plot by species
  mutate(Spp.Plot=Spp.Plot*20) %>%# plot multiplier to get sph

@@

- There are actually more spruce trees per hectare in the high-removal treatment than the other two.  
- Higher Q value for Control treatment SAF compared to harvested units.
-

@@

```{r}
dat %>% 
   left_join(plotDat,by="Plot") %>% # join with plot data
   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
  mutate(BA=(DBH/100/2)^2*pi) %>% # convert tree DBH to basal area
  group_by(BA.Target,Plot,Year) %>% 
  summarise(BA=sum(BA)*20) %>% # multiply by 20 for BA/ha
  ungroup() %>%

@@

@@

  # Plotting
  ggplot()+
  aes(x=Year,y=BA.Mean,color=label,group=label)+
  geom_point(size=3)+
  geom_line()+
  geom_errorbar(aes(ymin=BA.Mean-BA.se,ymax=BA.Mean+BA.se),width=0.1)+
  xlab("Measurement year")+
  ylab("Mean basal area per plot (m2/ha)")

@@

- Basal area actually decreased between 1992 and 1994 for control and Trt20.  I will look into this a bit more.
- Both 10 and 20 Treatment Units have converged in terms of basal area per hectare by 2019, and are >10m2/ha less than the control.

@@

  # Plotting
  ggplot()+
  aes(x=Year,y=BA.Mean,group=Species,fill=Species)+
  geom_bar(stat="Identity")+
  facet_wrap(~label)+
  ylab("Mean basal area per plot (m2/ha)")+
  xlab("Measurement year")

@@

- Spruce basal area increased relatively faster in the high-removal treatment unit compared to the other two treatment units.  Interesting!
- Douglas-fir basal area decreased between 1992 and 1994.  This likely explains the overall basal area decreases between 92 and 94 in previous figure.  Will look into this more  *Note - I removed dead trees for this analysis, just to see if this explained the decrease in Fdi BA - no effect*
- Close to 30 years afterwards, both harvest treatments are similar in terms of total basal area, and SAF area.

>>>>

## Prelim analysis
Below are some preliminary figures of the EP1162 data.

@@

NB2: Plots 16,17, 18 and 19 were omitted from the dataset for these analyses because they were not measured in 1994 or 1997.

```{r}
dat<-
  dat %>% 
  filter(!Plot %in% c(16,17,18,19)) # remove plots 16-19 from data for further analysis

This means that the following analyses are based on `r length(unique(dat$Plot))` plots.

@@

dat %>% 
  left_join(plotDat,by="Plot") %>% 
   filter(Status=="Live") %>% # filter for live trees only
  mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes

@@

### Trees per hectare by species and diameter class in 2019
Let's look at species composition by treatment unit in 2019:

@@

dat %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data
   filter(Status=="Live") %>% # filter for live trees only
   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
  filter(Year==2019) %>% # filter for 2019
  filter(Species%in%c("Sx","Bl")) %>% #filter for Sx and Bl
  group_by(BA.Target,Plot,Species,DBH.class) %>% 
  summarise(Spp.Plot=n()) %>% # summarise number of trees per plot by species
  mutate(Spp.Plot=Spp.Plot*20) %>%# plot multiplier to get sph

@@

- There are actually more spruce trees per hectare in the high-removal treatment than the other two.  
- Higher Q value for Control treatment SAF compared to harvested units.

### Basal area per hectare by species and diameter class in 2019
Let's look at basal area by species, diameter class and treatment unit in 2019:

```{r}
dat %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data
   filter(Status=="Live") %>% # filter for live trees only
   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
  filter(Year==2019) %>% # filter for 2019
  filter(Species%in%c("Sx","Bl")) %>% #filter for Sx and Bl
  mutate(BA=(DBH/2/100)^2*pi) %>% # convert DBH to basal area
  group_by(BA.Target,Plot,Species,DBH.class) %>% 
  summarise(BA.Plot=sum(BA)*20) %>%   # sum basal area per diam class and multiply to get per ha

  # create means
  ungroup() %>% 
  group_by(BA.Target,Species,DBH.class) %>%
  summarise(BA=mean(BA.Plot)) %>%

  # Plotting
  ggplot()+
  aes(x=DBH.class,y=BA,fill=Species)+
  geom_bar(stat="Identity")+
  facet_wrap(~label~Species,ncol=2)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1,size=6,vjust=0.25))+
  ylab("Mean basal area per hectare per plot (m2)")+
  xlab("Diameter classes (cm)")

- Sx basal area is concentrated in the larger diameter classes.
- Bl basal area is more evenly distributed across diameter classes.

@@

```{r}
dat %>% 
   left_join(plotDat,by="Plot") %>% # join with plot data
  filter(Status=="Live") %>% 
   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
  mutate(BA=(DBH/100/2)^2*pi) %>% # convert tree DBH to basal area
  group_by(BA.Target,Plot,Year) %>% 
  summarise(BA=sum(BA)*20) %>% # multiply by 20 for BA/ha
  ungroup() %>%

@@

    # Add labeller column
  mutate(Trt.Unit=factor(paste("Trt:",BA.Target,sep=""))) %>%

@@

  # Plotting
  ggplot()+
  aes(x=Year,y=BA.Mean,color=Trt.Unit,group=Trt.Unit)+
  geom_point(size=3)+
  geom_line()+
  geom_errorbar(aes(ymin=BA.Mean-BA.se,ymax=BA.Mean+BA.se),width=0.1)+
  xlab("Measurement year")+
  ylab("Mean basal area per plot (m2/ha)")

@@

- Both 10 and 20 Treatment Units have converged in terms of basal area per hectare by 2019, and are >10m2/ha less than the control.
- Treatment Unit 10 has a steeper temporal trend than Treatment Unit 20.  Growing faster?

@@

  # Plotting
  ggplot()+
  aes(x=Year,y=BA.Mean,fill=Species)+
  geom_bar(stat="Identity")+
  facet_wrap(~label)+
  ylab("Mean basal area per plot (m2/ha)")+
  xlab("Measurement year")

@@

- Spruce basal area increased relatively faster in the high-removal treatment unit compared to the other two treatment units.  Interesting!
- Close to 30 years afterwards, both harvest treatments are similar in terms of total basal area, and SAF area.
- 2019 BA in both treatment units are similar to control in 2009.  Treatment units are about 10 years behind at this point?

### PBAI
Below is code to recreate Mike's PBAI graph by initial BA:

dat %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data
   filter(Status=="Live") %>% # filter for live trees only
  mutate(BA=(DBH/2/100)^2*pi*20) %>% # convert DBH to basal area and multiply for per hectare
  dplyr::select(BA.Target,Plot,TreeID,Species,Year,BA) %>% 
  group_by(BA.Target,Plot,Year) %>% 
  summarise(BA.Sum=sum(BA,na.rm=T)) %>% 
  pivot_wider(names_from=Year,values_from=BA.Sum) %>% 
  mutate(`1992-1994`=`1994`-`1992`) %>% 
  mutate(`1994-1997`=`1997`-`1994`) %>% 
  mutate(`1997-2009`=`2009`-`1997`) %>% 
  mutate(`2009-2019`=`2019`-`2009`) %>%

  # pivot back
  dplyr::select(BA.Target,Plot,`1992-1994`:`2009-2019`) %>% 
  pivot_longer(cols=`1992-1994`:`2009-2019`,names_to="Period",values_to="PBAI") %>% 
  mutate(Period=factor(Period)) %>%

  # summarise PBAI by BA.Target
  group_by(BA.Target,Period) %>% 
  summarise(PBAI.mean=mean(PBAI,na.rm=T),
            PBAI.se=sd(PBAI,na.rm=T)/sqrt(length(PBAI))) %>%

    # Add labeller column
  mutate(Trt.Unit=factor(paste("Trt:",BA.Target,sep=""))) %>%

  # Plotting
  ggplot()+
  aes(x=Period,y=PBAI.mean,fill=Trt.Unit,color=Trt.Unit,group=Trt.Unit)+
  geom_point(size=3)+
  geom_line()+
  geom_errorbar(aes(ymin=PBAI.mean-PBAI.se,ymax=PBAI.mean+PBAI.se),width=0.1)+
  xlab("Period")+
  ylab("Mean basal area per plot (m2/ha)")

Graph looks similar to Mike's original figure.

Some very interesting trends here:

- Treatment 10 plots have higher increment than other treatment units.
- Control treatment plots increment is starting to stabilize now?  Does this suggest that in the next 10-20 years, the treatment units may converge with control in terms of BAI?

### Effect of initial BA on PBAI
Below is R code to re-create Mike's figure of initial BA on PBAI by period

pbai<- # assign output so we can plot several periods
  dat %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data
   filter(Status=="Live") %>% # filter for live trees only
  mutate(BA=(DBH/2/100)^2*pi*20) %>% # convert DBH to basal area and multiply for per hectare
  dplyr::select(BA.Target,Plot,TreeID,Species,Year,BA) %>% 
  group_by(BA.Target,Plot,Year) %>% 
  summarise(BA.Sum=sum(BA,na.rm=T)) %>% 
  pivot_wider(names_from=Year,values_from=BA.Sum) %>% 
  mutate(`1992-1994`=`1994`-`1992`) %>% 
  mutate(`1994-1997`=`1997`-`1994`) %>% 
  mutate(`1997-2009`=`2009`-`1997`) %>% 
  mutate(`2009-2019`=`2019`-`2009`)

  # Plotting function
pbai.Plot<-function(initYear,baPeriod,data=pbai) {

  p1<-
    data.frame(data[initYear],data[baPeriod]) %>% 
    setNames(c("initialBA","PBAI")) %>% 
     # filter(PBAI>=0) %>% 
      ggplot()+
    aes(x=initialBA,y=PBAI)+
    geom_point(size=3)+
    xlab(paste(initYear," plot basal area per hectare",sep=""))+
    ylab(paste("BAI per hectare for ",baPeriod, "Period"))

  return(p1)

}

pbai.Plot(initYear="2009",baPeriod="2009-2019")

We see a negative relationship between plot basal area and 2009-2019 PBAI.  Note the outlier - this plot is a control plot that had a large Sx tree die between 2009-2019, resulting in negative PBAI for that period.

The relationship is consistent using initial BA for earlier periods (e.g., 1994):

```{r}
pbai.Plot(initYear="1994",baPeriod="2009-2019")
```

*For discussion*  Although the relationship seems nonlinear negative, I think that's mostly due to the plot (Treatment: Control Plot 3) with negative PBAI.  I don't think we would want to model plots with negative PBAI, as these are due to stand dynamics/forest health factors, and we are primarily interested in effects of treatment on growth.

With plot 3 removed, the relationship looks more linear:

```{r}
pbai.2<-
  pbai %>% 
  filter(Plot!=3)

pbai.Plot(initYear="1994",baPeriod="2009-2019",data=pbai.2)

***
This linear relationship doesn't hold for periods earlier than the 2009-2019 period:

pbai.Plot(initYear="1994",baPeriod="1994-1997")
```
